---
title: "SLE 777 part 1 a4"
author: "Khoa Huynh"
date: "2025-09-23"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r}
library("R.utils")

URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL,destfile="gene_expression.tsv")

list.files()
```

```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
download.file(URL,destfile="growth_data.csv")
list.files()
```
# Question 1
#Read the file data, the first six genes


```{r}
gene_data <- read.delim("gene_expression.tsv", row.names = 1, header = TRUE)
head(gene_data, 6)
```
## Question 2
#Create new column with mean of other columns
```{r}
gene_data$Mean <- rowMeans(gene_data)
```
# Show a table of values for the first six genes
```{r}
head(gene_data, 6)
```
## Question 3
#List the 10 genes with the highest mean expression
```{r}
top10_genes <- gene_data[order(gene_data$Mean, decreasing = TRUE), ]
head(top10_genes, 10)
```
## Question 4
#Determine the number of genes with a mean <10
```{r}
num_genes_below_10 <- sum(gene_data$Mean < 10)
cat(num_genes_below_10)
```
#Question 5
#Make a histogram plot of the mean values
```{r}
hist(gene_data$Mean,
     breaks = 1000,
     main = "Distribution of Mean Gene Expression",
     xlab = "Mean Expression Level",
     ylab = "Frequency",
     xlim = c(0,30000),
     col = "blue")

```
# Question 6
#Import this growth_data file into an R object
```{r}
growth <- read.csv("growth_data.csv", header = TRUE)
```
##Column names of growth data
```{r}
colnames(growth)
```
## Question 7
#Calculate the mean and standard deviation of tree circumference at the start(2005) and end of the study (2020)

```{r}
# Mean and Standard Deviation (2005) & (2020)
growth_stats_raw <- aggregate(cbind(Circumf_2005_cm, Circumf_2020_cm) ~ Site,
                              data = growth,
                              FUN = function(x) c(mean = mean(x, na.rm = TRUE),
                                                  sd   = sd(x, na.rm = TRUE)))

# Split the matrix columns into separate columns
growth_stats_summary <- data.frame(
  Site        = growth_stats_raw$Site,
  Mean_2005   = growth_stats_raw$Circumf_2005_cm[, "mean"],
  SD_2005     = growth_stats_raw$Circumf_2005_cm[, "sd"],
  Mean_2020   = growth_stats_raw$Circumf_2020_cm[, "mean"],
  SD_2020     = growth_stats_raw$Circumf_2020_cm[, "sd"]
)

# Display table results
print(growth_stats_summary)

```
# Question 8
#Make a box plot of tree circumference at the start and end of the study at both sites.
```{r}

# Prepare data for boxplot
northeast_start <- growth$Circumf_2005_cm[growth$Site == "northeast"]
northeast_end <- growth$Circumf_2020_cm[growth$Site == "northeast"]
southwest_start <- growth$Circumf_2005_cm[growth$Site == "southwest"]
southwest_end <- growth$Circumf_2020_cm[growth$Site == "southwest"]

# Create boxplot
boxplot(northeast_start, northeast_end, southwest_start, southwest_end,
        names = c("Northeast\nStart", "Northeast\nEnd", "Southwest\nStart", "Southwest\nEnd"),
        main = "Tree Circumference by Site and Time Point",
        xlab = "Site and Time Point",
        ylab = "Circumference (cm)",
        col = c("lightcoral", "blue", "lightcoral", "blue"),
        border = "black")

# Add a legend
legend("topleft", 
       legend = c("Start", "End"), 
       fill = c("lightcoral", "blue"),
       title = "Time Point")
```

# question 9 
##Calculate the mean growth over the last 10 years at each site.
```{r}
# Growth from 2010 to 2020
growth$Growth_10yr <- growth$Circumf_2020_cm - growth$Circumf_2010_cm
# Mean 10 year growth for each site
mean_growth_by_site <- aggregate(Growth_10yr ~ Site, data = growth, FUN = mean)

# Show results
print(mean_growth_by_site)

```
# question 10 
#T-test to compare 10-year growth between sites
```{r}

t_test_result <- t.test(Growth_10yr ~ Site, data = growth)

## Print t-test results with interpretation
cat("\n T-Test: Comparing 10-Year Growth Between Sites\n")
cat("\n T-Test: Comparing 10-Year Growth Between Sites\n")
cat("P-value:", round(t_test_result$p.value, 4), "\n")
cat("95% Confidence Interval:", 
    round(t_test_result$conf.int[1], 2), "to", 
    round(t_test_result$conf.int[2], 2), "\n")
cat("Mean Growth - Northeast:", round(t_test_result$estimate["mean in group Northeast"], 2), "cm\n")
cat("Mean Growth - Southwest:", round(t_test_result$estimate["mean in group Southwest"], 2), "cm\n")

# Interpretation
if (t_test_result$p.value < 0.05) {
  cat("Conclusion: There is a statistically significant difference in 10-year growth between sites (p < 0.05).\n")
} else {
  cat(" Conclusion: No statistically significant difference in 10-year growth between sites (p ≥ 0.05).\n")
}

```



